---
title: "DESeq-SENvsRES"
output: html_document
date: "2023-08-22"
---

#LOADING LIBRARIES
```{r}
suppressMessages(library(DESeq2))
suppressMessages(library(ggplot2))
suppressMessages(library(gplots))
suppressMessages(library(plotly))
suppressMessages(library(RColorBrewer))
library(ggvenn)
library(sva)
```


#SET WORKING DIRECTORY
```{r}
rm(list = ls())
setwd('/Users/amyhayward/Documents/BIOINFORMATICS MSC/Project/FINAL_FILES')
```

#LOAD SAMPLE SHEET
```{r}
sample_sheet_file <- 'OLA_TERTILE_LN.csv'
sample_sheet <- read.table(sample_sheet_file, header = T, sep = ',', row.names = 1)
sample_sheet$resistance_label <- gsub("[^A-Za-z0-9_.]", "_", sample_sheet$resistance_label)
```

#SUBSET SAMPLE SHEET TO MORE RESISTANT AND LESS RESISTANT
```{r}
sample_sheet <- sample_sheet[sample_sheet$resistance_label %in% c("less_resistant", "more_resistant"), ]
```

#LOAD FILE SO WE CAN MATCH SAMPLES TO TRAINING DATA
```{r}
cols_to_match_file <- "X_TRAIN_MATCH_OLA.csv"
cols_to_match <- read.table(cols_to_match_file, header = T, sep = ',', row.names = 1, check.names = F)
```

#FIND INTERSECTION BETWEEN SAMPLE SHEET AND TRAINING DATA
```{r}
intersection <- intersect(rownames(sample_sheet), rownames(cols_to_match))

sample_sheet <- sample_sheet[intersection, ]
cols_to_match <- cols_to_match[intersection, ]
```


#LOAD EXPRESSION DATA
```{r}
raw_counts_file <- "RAW_CLEAN_DESEQ_OLA.csv"
raw_counts <- read.table(raw_counts_file, header = T, sep = ',', row.names = 1, check.names = F)
```


```{r}
intersect2 <- intersect(names(raw_counts), names(cols_to_match))
raw_counts <- raw_counts[,intersect2]

intersect3 <- intersect(rownames(raw_counts), rownames(sample_sheet))
raw_counts <- raw_counts[intersect3,]

raw_counts[raw_counts == 0] <- 1 #add pseudo-count to zeros for deseq
```


```{r}
head(raw_counts)
```

#MATCH ROW ORDER OF SAMPLE SHEET AND COUNT DATA
```{r}
order <- rownames(sample_sheet)
raw_counts <- raw_counts[order, ]
```

#MATCH SAMPLE ORDER AND CREATE MATRIX
```{r}
raw_matrix <- as.matrix(raw_counts)
raw_matrix <- t(raw_matrix)
```

```{r}
dim(raw_matrix)
```


#APPLY DESeq

```{r}
dds <- DESeqDataSetFromMatrix(raw_matrix, colData = sample_sheet, design = ~resistance_label)
```


```{r}
dds <- DESeq(dds)
plotDispEsts(dds)
```

#GET RESULTS NAME
```{r}
# Access the results for a specific comparison
resultsNames(dds)[-1]
```

#RESULTS TABLE
```{r}
res_less_vs_more <- results(dds, name = "resistance_label_more_resistant_vs_less_resistant")
```

#GET RESULTS WHICH PVALUE <0.05 AND PADJ <0.1
```{r}
cat(paste0('significant p-values = ', sum(res_less_vs_more$pvalue < 0.05, na.rm = T), '\n'))
cat(paste0('significant p-adjust = ', sum(res_less_vs_more$padj < 0.05, na.rm = T), '\n'))
```


#OBTAIN NORMALISED READ COUNTS
```{r}
norm_counts <- counts(dds) 
```

#OBTAIN VST TRANSFORMED COUNTS
```{r}
vds <- varianceStabilizingTransformation(dds)
vst_counts <- as.matrix(assay(vds))
```


#RESULTS VISUALISATION
```{r}
pca.plot <- function(read.counts, classes, 
                     comps = c(1, 2), ntop = min(500, nrow(read.counts)), standard = T,
                     col = c('lightblue', 'orange', 'MediumVioletRed', 'SpringGreen')){
  top_index <- order(rowVars(read.counts), decreasing = TRUE)[1:ntop]
  pca <- prcomp(scale(t(read.counts[top_index,]), center = standard, scale = standard))
  prp <- pca$sdev^2 * 100 / sum(pca$sdev^2)
  
  pca_comps <- pca$x[,comps]
  prp_comps <- round(prp[comps], 2)

  df <- data.frame(pc1 = pca_comps[,1], pc2 = pca_comps[,2], condition = classes)
  p  <- ggplot(df, aes(x = pc1, y = pc2, color = condition)) + 
        geom_point(size = 3) + 
        labs(title = paste0('Principal Component Analysis - PCs ', comps[1] , ' and ', comps[2]), 
             x = paste0('PC', comps[1], ' (', prp_comps[1], '%)'), 
             y = paste0('PC', comps[2], ' (', prp_comps[2], '%)')) + 
        geom_text(label = colnames(read.counts)) +
        scale_color_manual(values = col)
  return(p)
}
```

#PLOT PCA

```{r}

conds <- as.factor(sample_sheet$resistance_label)

p <- pca.plot(vst_counts, conds, comps = c(1,2), ntop =50, 
              col = brewer.pal(max(3, nlevels(conds)), 'Set2'))
ggplotly(p)
```


#RESEARCH QVALUE
```{r}
p <- ggplot(as.data.frame(res_less_vs_more), aes(x = pvalue)) + geom_histogram(fill = '#619CFF', bins = 100)
p

```

#PLOT MA
```{r}
plotMA(res_less_vs_more)
```


#PLOT HEATMAP
```{r}
png(filename="OLA-WRITEUP.png", width=900, bg="white")
par(mar=c(2,4,4,10)+.1)
heatmap.2(
  t(log10(vst_counts+1)[which(res_less_vs_more$padj < 0.05),]), trace='none', cexRow=1.5,cexCol=1,margins=c(12,8))
graphics.off()
```


#VOLCANO PLOT
```{r}
valid_index <- which(!is.na(res_less_vs_more$padj))
    df <- data.frame(
      pvals = res_less_vs_more$padj[valid_index],
      lfcs = res_less_vs_more$log2FoldChange[valid_index],
      gene = rownames(res_less_vs_more)[valid_index])
  
df$significant = as.factor(df$pvals < 0.05)
    
p <- ggplot(df, aes(x = lfcs, y = -log10(pvals), colour = significant, gene = gene)) +
      geom_point(alpha = 0.8, size = 2) +
      labs(title = 'Volcano Plot, Olaparib, Less Resistant vs Intermediate',x = 'log2 fold change', y = '-log10 p-value') +
      scale_color_manual(values = c('#619CFF', '#F564E3'))
      p
```


#SAVE RESULTS
```{r}
genes <- sort(row.names(dds))
res.table <- DataFrame(row.names = genes, log2FoldChange = res_less_vs_more$log2FoldChange, pvalue = res_less_vs_more$pvalue, padj = res_less_vs_more$padj)

write.csv(res.table, file = paste0('DESeq2_results-ola-res_more_vs_less_res.csv'))
```