---
title: "DESeq_TALA"
output: html_document
date: "2023-07-27"
---

#LOADING LIBRARIES
```{r}
suppressMessages(library(DESeq2))
suppressMessages(library(ggplot2))
suppressMessages(library(gplots))
suppressMessages(library(plotly))
suppressMessages(library(RColorBrewer))
library(ggvenn)
library(sva)
```

#SET WORKING DIRECTORY
```{r}
rm(list = ls())
setwd('/Users/amyhayward/Documents/BIOINFORMATICS MSC/Project/FINAL_FILES')
```

#LOAD SAMPLE SHEET
```{r}
sample_sheet_file <- 'TALA_TERTILE_LN.csv'
sample_sheet <- read.table(sample_sheet_file, header = T, sep = ',', row.names = 1)
sample_sheet$resistance_label <- gsub("[^A-Za-z0-9_.]", "_", sample_sheet$resistance_label)
```

#LOAD EXPRESSION DATA
```{r}
raw_counts_file <- "RAW_CLEAN_DESEQ_TALA.csv"
raw_counts <- read.table(raw_counts_file, header = T, sep = ',', row.names = 1, check.names = F)
```


#LOAD FILE SO WE CAN MATCH SAMPLES TO TRAINING DATA
```{r}
cols_to_match_file <- "X_TRAIN_MATCH_TALA.csv"
cols_to_match <- read.table(cols_to_match_file, header = T, sep = ',', row.names = 1, check.names = F)

dup_cols <- intersect(names(raw_counts), names(cols_to_match))
raw_counts <- raw_counts[,dup_cols]
raw_counts[raw_counts == 0] <- 1
```

```{r}
head(raw_counts)
```

#MATCH SAMPLE ORDER AND CREATE MATRIX
```{r}
sample_sheet <- sample_sheet[rownames(raw_counts), ]

raw_matrix <- as.matrix(raw_counts)
raw_matrix <- t(raw_matrix)
```

```{r}
raw_matrix[raw_matrix == 0] <- 1
```

```{r}
dim(raw_matrix)
```


#APPLY DESeq

```{r}
dds <- DESeqDataSetFromMatrix(raw_matrix, colData = sample_sheet, design = ~resistance_label)
```

#DISPERSION PLOT
```{r}
dds <- DESeq(dds)
plotDispEsts(dds)
```

#GET RESULTS NAME
```{r}
resultsNames(dds)[-1]
```

```{r}
res_less_vs_int <- results(dds, name = "resistance_label_less_resistant_vs_intermediate")
res_more_vs_int <- results(dds, name = "resistance_label_more_resistant_vs_intermediate")
```


#GET RESULTS WHICH PVALUE <0.05 AND PADJ <0.1
```{r}
cat(paste0('significant p-values = ', sum(res_less_vs_int$pvalue < 0.05, na.rm = T), '\n'))
cat(paste0('significant p-adjust = ', sum(res_less_vs_int$padj < 0.05, na.rm = T), '\n'))
```

```{r}
cat(paste0('significant p-values = ', sum(res_more_vs_int$pvalue < 0.05, na.rm = T), '\n'))
cat(paste0('significant p-adjust = ', sum(res_more_vs_int$padj < 0.05, na.rm = T), '\n'))
```


#OBTAIN NORMALISED READ COUNTS
```{r}
norm_counts <- counts(dds) 
```

#OBTAIN VST TRANSFORMED COUNTS
```{r}
vds <- varianceStabilizingTransformation(dds)
vst_counts <- as.matrix(assay(vds))
```

#RESULTS VISUALISATION
```{r}
pca.plot <- function(read.counts, classes, 
                     comps = c(1, 2), ntop = min(500, nrow(read.counts)), standard = T,
                     col = c('lightblue', 'orange', 'MediumVioletRed', 'SpringGreen')){
  top_index <- order(rowVars(read.counts), decreasing = TRUE)[1:ntop]
  pca <- prcomp(scale(t(read.counts[top_index,]), center = standard, scale = standard))
  prp <- pca$sdev^2 * 100 / sum(pca$sdev^2)
  
  pca_comps <- pca$x[,comps]
  prp_comps <- round(prp[comps], 2)

  df <- data.frame(pc1 = pca_comps[,1], pc2 = pca_comps[,2], condition = classes)
  p  <- ggplot(df, aes(x = pc1, y = pc2, color = condition)) + 
        geom_point(size = 3) + 
        labs(title = paste0('Principal Component Analysis - PCs ', comps[1] , ' and ', comps[2]), 
             x = paste0('PC', comps[1], ' (', prp_comps[1], '%)'), 
             y = paste0('PC', comps[2], ' (', prp_comps[2], '%)')) + 
        geom_text(label = colnames(read.counts)) +
        scale_color_manual(values = col)
  return(p)
}
```

#PLOT PCA

```{r}

conds <- as.factor(sample_sheet$resistance_label)

p <- pca.plot(vst_counts, conds, comps = c(1,2), ntop =50, 
              col = brewer.pal(max(3, nlevels(conds)), 'Set2'))
ggplotly(p)
```



#RESEARCH QVALUE
```{r}
p <- ggplot(as.data.frame(res_less_vs_int), aes(x = pvalue)) + geom_histogram(fill = '#619CFF', bins = 100)
p

```
```{r}
p <- ggplot(as.data.frame(res_more_vs_int), aes(x = pvalue)) + geom_histogram(fill = '#619CFF', bins = 100)
p
```

#PLOT MA
```{r}
plotMA(res_less_vs_int)
```
```{r}
plotMA(res_more_vs_int)
```

#PLOT HEATMAP
```{r}
png(filename="TALA_LESS_INT.png", width=900, bg="white")
par(mar=c(2,4,4,10)+.1)
heatmap.2(
  t(log10(vst_counts+1)[which(res_less_vs_int$padj < 0.05),]), trace='none', cexRow=1.5,cexCol=1,margins=c(12,8))
graphics.off()

```

```{r}
png(filename="TALA_MORE_INT.png", width=900, bg="white")
par(mar=c(2,4,4,10)+.1)
heatmap.2(
  t(log10(vst_counts+1)[which(res_more_vs_int$padj < 0.05),]), trace='none', cexRow=1.5,cexCol=1,margins=c(12,8))
graphics.off()
```


#VOLCANO PLOT
```{r}
valid_index <- which(!is.na(res_less_vs_int$padj))
    df <- data.frame(
      pvals = res_less_vs_int$padj[valid_index],
      lfcs = res_less_vs_int$log2FoldChange[valid_index],
      gene = rownames(res_less_vs_int)[valid_index])
  
df$significant = as.factor(df$pvals < 0.05)
    
p <- ggplot(df, aes(x = lfcs, y = -log10(pvals), colour = significant, gene = gene)) +
      geom_point(alpha = 0.8, size = 2) +
      labs(title = 'Volcano Plot, Talazoparib, Less Resistant vs Intermediate',x = 'log2 fold change', y = '-log10 p-value') +
      scale_color_manual(values = c('#619CFF', '#F564E3'))
      p
```
```{r}

#VOLCANO PLOT

valid_index <- which(!is.na(res_more_vs_int$padj))
    df <- data.frame(
      pvals = res_more_vs_int$padj[valid_index],
      lfcs = res_more_vs_int$log2FoldChange[valid_index],
      gene = rownames(res_more_vs_int)[valid_index])
  
df$significant = as.factor(df$pvals < 0.05)
    
p <- ggplot(df, aes(x = lfcs, y = -log10(pvals), colour = significant, gene = gene)) +
      geom_point(alpha = 0.8, size = 2) +
      labs(title = 'Volcano Plot, Talazoparib, More Resistant vs Intermediate',x = 'log2 fold change', y = '-log10 p-value') +
      scale_color_manual(values = c('#619CFF', '#F564E3'))
      p
```




#SAVE RESULTS
```{r}
genes <- sort(row.names(dds))
res.table <- DataFrame(row.names = genes, log2FoldChange = res_more_vs_int$log2FoldChange, pvalue = res_more_vs_int$pvalue, padj = res_more_vs_int$padj)

write.csv(res.table, file = paste0('DESeq2_results-tala-res_more_vs_int.csv'))
```

```{r}
genes <- sort(row.names(dds))
res.table <- DataFrame(row.names = genes, log2FoldChange = res_less_vs_int$log2FoldChange, pvalue = res_less_vs_int$pvalue, padj = res_less_vs_int$padj)

write.csv(res.table, file = paste0('DESeq2_results-tala-res_less_vs_int.csv'))
```


```{r}
source <- '~/Documents/BIOINFORMATICS MSC/Project/FINAL_FILES/6(TALA)DESEQ_CODE_PROJECT_LESS-INT-MORE.Rmd'
new <- '~/Documents/BIOINFORMATICS MSC/Project/FINAL_FILES/6(OLA)DESEQ_CODE_PROJECT_LESS-INT-MORE.Rmd'

file.copy(source, new)
```

